Informing policy via dynamic models: Cholera in Haiti

### 2023-09-28 ms.Rnw
Finished addressing Anna's + Kevin's comments

\section*{Author summary}
Quantitative understanding of infectious disease transmission dynamics relies upon mathematical models informed by scientific knowledge and relevant data.
The models aim to provide a statistical description of the trajectory of an epidemic and its uncertainty, together with a representation of the underlying biological mechanisms.
Evaluation of success at these goals is necessary in order for a model to provide a reliable tool for guiding evidence-based public policy interventions.
In this article, we discuss common modeling decisions that may not provide an adequate statistical explanation of the data, and we show that attention to statistical fit can lead to more accurate policy evaluations.
We do this by conducting a re-analysis of the 2012-2019 cholera outbreak in Haiti.
Our analysis presents methodology for diagnosing how well a model describes observed data, and provides careful justification of valid conclusions from the fitted model.
Objective measures are used to benchmark model fit; when these are combined with reproducibility, a framework emerges for continual improvment when revisiting the data and models.
Our data analysis workflow is supported by recent advances in algorithms, software and hardware, which facilitate statistical fitting of nonlinear stochastic dynamic models to observed incidence data.
However, inference for high-dimensional systems remains a methodological challenge.
One of the models under consideration involves spatially coupled stochastic meta-populations, and we demonstrate how a recently developed algorithm permits likelihood-based inference and model diagnostics in this setting.
We contend that raising the currently accepted standards of infectious disease modeling will result in a greater ability of scientists and policy makers to understand and respond to future infectious disease outbreaks.

@@

The four independent teams were given the task of estimating the
potential effect of prospective oral cholera vaccine (OCV) programs.
While OCV is accepted as a safe and effective tool for controlling the spread of cholera, the global stockpile of OCV doses remains limited \cite{pezzoli20}.
Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program \cite{lee20}.
In the study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
The groups largely adhered to existing guidelines for creating models to inform policy \cite{behrend20,saltelli20} and, despite their autonomy, obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
Their forecasts, however, were inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.
Though cholera has recently reemerged in Haiti \cite{rubin22}, the inability to accurately forecast cholera incidence from 2019-2022---while diligently adhering to existing scientific standards for infectious disease modeling---demonstrates that existing standards of disease modeling and forecasting may be insufficient.

@@

While this work is focused on the models proposed by Lee et al.~\cite{lee20}, our suggestions have broader relevance.
To investigate the extent to which \cite{lee20} is typical of the substantial body of work on Haiti cholera dynamics, we preformed a literature review by searching PubMed with keywords: Haiti, cholera, model.
The search resulted in 66 papers, of which 32 used dynamic models to describe the cholera epidemic in Haiti.
The models make various choices on the dichotomies considered by Lee et al.~\cite{lee20}; deterministic versus stochastic; compartment model versus agent-based; aggregated versus spatially explicit.

@@

\multirow{2}{*}{Birth/death (yr)}
  & ${\muBirth^{-1} = \Sexpr{myround(1/p1["mu"]/52,1)}}\fixed$

@@

  & ${\muDeath^{-1} = \Sexpr{myround(1/p3["mu1"],1)}}\fixed$

@@

& ${\muDeath^{-1} = \Sexpr{myround(1/p1["delta"]/52,1)}}\fixed$

@@

& ${\choleraDeath^{-1} = \Sexpr{myround(1/p3["alpha1"],2)}}\fixed$

@@

Parameters that were fixed and not calibrated using the data are indicated with $\fixed$;
parameters that were added during our re-analysis and were not considered by Lee et al. are indicated with $^*$.

@@

In this article we calibrate the parameters of each of our models by maximizing the likelihood, as described in Eq.~\myeqref{eq:likedef}.
The likelihood for each of the fitted models---and the corresponding AIC values for model comparisons that include an adjustment for the number of calibrated parameters---is provided in Table~\ref{tab:likes}
In the following subsections we describe in detail our approach to calibrating the three proposed mechanistic models to observed cholera incidence data.

@@

One of the most frequently used approaches to calibrate the parameters of a statistical model is maximum likelihood estimation.
In the case of a POMP model, the likelihood function---as described in Eq.~\myeqref{eq:likedef}---is written as $\mathcal{L}(\paramVec) = \int f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big) d\bm{x}_{0:N}$.
The high-dimensional integral in the likelihood, and the potentially non-linear nature of the process model, make it impossible to obtain a closed form expression of the likelihood function of POMP models except in the most simple cases.
The process model that defines Model~1 was developed with the intent of providing a plausible description of cholera dynamics rather than a providing statistically convenient description.
Because of this, a closed form expression of the likelihood of Model~1 is not readily available.

In order to retain the ability to propose models that are scientifically meaningful rather than only those that are simply statistically convenient, we restrict ourselves to parameter estimation techniques that have the plug-and-play property, which is that the fitting procedure only requires the ability to simulate the latent process instead of evaluating transition densities \cite{breto09,he10};
in the context of the notation and definitions employed in this article, this means that we only require the ability to simulate from $f_{\bm{X}_0}\big(\bm{x}_0;\paramVec\big)$ and $f_{\bm{X}_n|\bm{X}_{n-1}}\big(\bm{x}_{n}|\bm{x}_{n-1}; \paramVec\big)$ rather than needing to evaluate these densities.
Plug-and-play algorithms include Bayesian approaches like ABC and PMCMC \cite{toni09,andrieu10}, but here we use algorithms that enable maximum likelihood estimation.
To our knowledge, the only plug-and-play methods that can be used to maximize the likelihood for POMP models of this complexity are iterated filtering algorithms \cite{ionides15}, which modify the well-known \emph{particle filter} \cite{arulampalam02}.
The particle filter is often used to obtain values of latent states, but can also be used to obtain likelihood estimates of a POMP model.
Iterated filtering algorithms such as IF2 \cite{ionides15} extend the particle filter by performing a random walk for each parameter and particle;
these perturbations are carried out iteratively over multiple filtering operations, using the collection of parameters from the previous filtering pass as the parameter initialization for the next iteration, and decreasing the random walk variance at each step.

@@

@@

Another approach is to compare a quantitative measure of the model fit (such as MSE, predictive accuracy, or model likelihood) among all proposed models.
These comparisons, which provide insight into how each model performs relative to the others, are quite common \cite{rinaldo12,sallah17}.
To calibrate relative measures of fit, it is useful to compare against a model that has well-understood statistical ability to fit data, and we call this model a {\it benchmark}.
Standard statistical models, interpreted as associative models without requiring any mechanistic interpretation of their parameters, provide suitable benchmarks.
Examples include linear regression, auto-regressive moving average time series models, or even independent and identically distributed measurements.
Benchmarks enable us to evaluate the goodness of fit that can be expected of a suitable mechanistic model.

@@

In some cases, a possible benchmark model could be a generally accepted mechanistic model, but often no such model is available.
Because of this, we use a simple negative binomial model with an auto regressive mean as our associative benchmark; this model is described in \myeqref{eq:AR_NB}.
\begin{eqnarray}
\label{eq:AR_NB}
Y_n|Y_{n - 1} \sim \text{NB}\big(\NBintercept + \NBar Y_{n-1}, \NBsize \big),
\end{eqnarray}
where $\mathrm{E}\big(Y_n|Y_{n - 1}\big) = \NBintercept + \NBar Y_{n-1}$, and $\Var\big(Y_n|Y_{n - 1}\big) = \mathrm{E}\big(Y_n|Y_{n - 1}\big) + \mathrm{E}\big(Y_n|Y_{n - 1}\big)^2 / \NBsize$.
In the case where the case counts are large, an alternative benchmark recommended by He et al.~\cite{he10} is a log-linear Gaussian ARMA model;
the theory and practice of ARMA models is well developed, and these linear models are appropriate on a log scale due to the exponential growth and decay characteristic of biological dynamics.
We chose to use the auto-regressive negative binomial model, however, because the large number of weeks with zero recorded cholera cases in department level data makes a benchmark based on a continuous distribution problematic.
Likelihoods of Models~1--3 and their respective ARMA benchmark models are provided in Table~\ref{tab:likes}.

@@

Our case study provides an example of this in the parameter estimate for the duration of natural immunity due to cholera infection, $\muRS^{-1}$.
Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],3)} \, \mathrm{yr}$, suggesting that individuals have effectively permanent immunity to cholera once infected.
To interpret this result, we bear in mind that the data ranged from 2010-2019, and therefore estimates of immunity longer than $10\, \mathrm{yr}$---the upper end of previous estimates of natural immunity \cite{king08}---effectively result in the same model dynamics.
The depletion of susceptible individuals may also be attributed to confounding mechanisms---such as localized vaccination programs and non-pharmaceutical interventions that reduce cholera transmission \cite{trevisin22, rebaudet21}---that were not accounted for in the model.
Perhaps the best interpretation of the estimated parameter, then, is that under model assumptions, the model most adequately describes the observed data by having a steady decrease in the number of susceptible individuals.
The weak statistical fit of Model~2 compared to a log-linear benchmark (see Table~\ref{tab:likes}) cautions us against drawing quantitative conclusions from this model.
A model that has a poor statistical fit may nevertheless provide a useful conceptual framework for thinking about the system under investigation.
However, a claim that the model has been validated against data should be reserved for situations where the model provides a statistical fit that is competitive against alternative explanations.

\subsection*{Robust interpretation of model based conclusions}\label{sec:advice}

In this section, we make some general points on deriving scientific conclusions from statistical analysis of mechanistic models, and connect these points to our specific case study.

@@

If a mechanistic model including a feature (such as a representation of a mechanism, or the inclusion of a covariate) fits better than mechanistic models without that feature, and also has competitive fit compared to associative benchmarks, this may be taken as evidence supporting the scientific relevance of the feature.
As for any analysis of observational data, we must be alert to the possibility of confounding.
For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled or unmeasured covariate.
For a mechanism, the model feature could in principle explain the data by helping to account for some different unmodeled phenomenon.
In the context of our analysis, the statistical evidence of a trend in transmission rate in Model~1 could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from colinear covariates.
Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate, resulting in confounded mechanisms.
The robust statistical conclusion is that a model which allows for change fits better than one which does not---we argue that a decreasing transmission rate is a plausible way to explain this, but the incidence data themselves do not provide enough information to pin down the mechanism.

@@

Recent information about a dynamic system should be more relevant for a forecast than older information.
This assertion may seem self-evident, but it is not the case for deterministic models, for which the initial conditions together with the parameters are sufficient for forecasting, and so recent data do not have special importance.
Epidemiological forecasts based on deterministic models are not uncommon in practice, despite their limitations \cite{king15}.
That may explain why Lee et al.~\cite{lee20} chose to obtain forecasts by simulating the calibrated models forward from initial conditions.
Here, we obtain forecasts by simulating future values using latent states that are consistent with the most recent observations by drawing latent states at time the last observation time $t_N$ from the filtering distribution $f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$,
where $\hat\paramVec$ is a vector of calibrated parameters.
The decision not to do this partially explains the unsuccessful forecasts of Lee et al.~\cite{lee20}: their Table~S7 shows that the subset of their simulations which were consistent with observing zero cases in 2019 also accurately predicted the prolonged absence of detected cholera.

>>>>

\section*{Author summary}
Quantitative understanding of infectious disease transmission dynamics relies upon mathematical models informed by scientific knowledge and relevant data.
The models aim to provide a statistical description of the trajectory of an epidemic and its uncertainty, together with a representation of the underlying biological mechanisms.
Evaluation of success at these goals is necessary in order for a model to provide a reliable tool for guiding evidence-based public policy interventions.
In this article, we discuss common modeling decisions that may not provide an adequate statistical explanation of the data, and we show that attention to statistical fit can lead to more accurate policy evaluations.
We do this by conducting a re-analysis of the 2010 cholera outbreak in Haiti.
Our analysis presents methodology for diagnosing how well a model describes observed data, and provides careful justification of valid conclusions from the fitted model.
Objective measures are used to benchmark model fit; when these are combined with reproducibility, a framework emerges for continual improvement when revisiting the data and models.
Our data analysis workflow is supported by recent advances in algorithms, software and hardware, which facilitate statistical fitting of nonlinear stochastic dynamic models to observed incidence data.
However, inference for high-dimensional systems remains a methodological challenge.
One of the models under consideration involves spatially coupled stochastic meta-populations, and we demonstrate how a recently developed algorithm permits likelihood-based inference and model diagnostics in this setting.
We contend that raising the currently accepted standards of infectious disease modeling will result in a greater ability of scientists and policy makers to understand and respond to future infectious disease outbreaks.

@@

The four independent teams were given the task of estimating the
potential effect of prospective oral cholera vaccine (OCV) programs.
While OCV is accepted as a safe and effective tool for controlling the spread of cholera, the global stockpile of OCV doses remains limited \cite{pezzoli20}.
Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program \cite{lee20}.
In the study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
The groups largely adhered to existing guidelines for creating models to inform policy \cite{behrend20,saltelli20} and, despite their autonomy, obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
Their forecasts, however, were inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.
Though cholera has recently reemerged in Haiti \cite{rubin22,PAHO23}, the inability to accurately forecast cholera incidence from 2019-2022---while diligently adhering to existing scientific standards for infectious disease modeling---demonstrates that existing standards of disease modeling and forecasting may be insufficient.

@@

While this work is focused on the models proposed by Lee et al.~\cite{lee20}, our suggestions have broader relevance.
To investigate the extent to which \cite{lee20} is typical of the substantial body of work on Haiti cholera dynamics, we performed a literature review by searching PubMed with keywords: Haiti, cholera, model.
The search resulted in 66 papers, of which 32 used dynamic models to describe the cholera epidemic in Haiti.
The models make various choices on the dichotomies considered by Lee et al.~\cite{lee20}; deterministic versus stochastic; compartment model versus agent-based; aggregated versus spatially explicit.

@@

\multirow{2}{*}{Birth/death (yr\textsuperscript{-1})}
  & ${\muBirth =  10^{-2}\times \Sexpr{myround(100 * 52 * p1["mu"],2)}}\fixed$

@@

  & ${\muDeath = 10^{-2}\times\Sexpr{myround(100 * p3["mu1"],2)}}\fixed$

@@

& ${\muDeath = 10^{-3}\times\Sexpr{myround(1000*52*p1["delta"],1)}}\fixed$

@@

& ${\choleraDeath = \Sexpr{myround(p3["alpha1"],2)}}\fixed$

@@

Parameters that were fixed and not calibrated using the data are indicated with $\fixed$; all fixed parameters values were chosen to match the fixed parameter values of \cite{lee20}.
Parameters that were added during our re-analysis and were not considered by Lee et al. are indicated with $^*$.

@@

In this article we calibrate the parameters of each of our models by maximizing the likelihood, as described in Eq.~\myeqref{eq:likedef}.
The likelihood for each of the fitted models---and the corresponding AIC values for model comparisons that include an adjustment for the number of calibrated parameters---is provided in Table~\ref{tab:likes}.
In the following subsections we describe in detail our approach to calibrating the three proposed mechanistic models to observed cholera incidence data.

@@

Model~1 was developed with the intent of providing a description of cholera dynamics that is scientifically plausible rather than one that is statistically convenient.
This results in the inability to obtain a closed form expression of the non-linear joint model density---described in Eq.~\myeqref{eq:jointLik}.
Therefore in order to perform likelihood based inference on this model, we are restricted to use parameter estimation techniques that have the \emph{plug-and-play} property, which is that the fitting procedure only requires the ability to simulate the latent process rather than evaluating transition densities \cite{breto09,he10};
in the context of the notation and definitions employed in this article, this means that we only require the ability to simulate from $f_{\bm{X}_0}\big(\bm{x}_0;\paramVec\big)$ and $f_{\bm{X}_n|\bm{X}_{n-1}}\big(\bm{x}_{n}|\bm{x}_{n-1}; \paramVec\big)$ rather than needing to evaluate these densities.
Plug-and-play algorithms include Bayesian approaches like ABC and PMCMC \cite{toni09,andrieu10}, but here we use algorithms that enable maximum likelihood estimation.
To our knowledge, the only plug-and-play methods that can be used to maximize the likelihood for POMP models of this complexity are iterated filtering algorithms \cite{ionides15}, which modify the well-known \emph{particle filter} \cite{arulampalam02}.

The particle filter, also referred to as sequential Monte Carlo, is a simulation based method that is frequently used in Bayesian inference to approximate the posterior distribution of latent states.
This algorithm can also be used to accurately approximate the log-likelihood of a POMP model, defined as the integral in Eq.~\myeqref{eq:likedef}.
Iterated filtering---algorithms such as IF2 \cite{ionides15}---extend the particle filter by performing a random walk for each parameter and particle;
these perturbations are carried out iteratively over multiple filtering operations, using the collection of parameters from the previous filtering pass as the parameter initialization for the next iteration, and decreasing the random walk variance at each step.
With a sufficient number of iterations, the resulting parameter values converge to a region of the parameter space that maximizes the model likelihood.

% One of the most frequently used approaches to calibrate the parameters of a statistical model is maximum likelihood estimation.
% In particular, a closed form expression of the joint model density---described in Eq.~\myeqref{eq:jointLik}---is not readily available, making likelihood based inference a challenging task.
% In the case of any POMP model, the likelihood function---as described in Eq.~\myeqref{eq:likedef}---is written as $\mathcal{L}(\paramVec) = \int f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big) d\bm{x}_{0:N}$.
% The high-dimensional integral in the likelihood, and the potentially non-linear nature of the process model, make it impossible to obtain a closed form expression of the likelihood function of POMP models except in the most simple cases.
% Model~1 was developed with the intent of providing a description of cholera dynamics that is scientifically plausible rather than on that is simply statistically convenient.
% Because of this, a closed form expression of the likelihood of Model~1 is not readily available.

% In order to retain the ability to propose models that are scientifically meaningful rather than only those that are simply statistically convenient, we restrict ourselves to parameter estimation techniques that have the plug-and-play property, which is that the fitting procedure only requires the ability to simulate the latent process instead of evaluating transition densities \cite{breto09,he10};
% in the context of the notation and definitions employed in this article, this means that we only require the ability to simulate from $f_{\bm{X}_0}\big(\bm{x}_0;\paramVec\big)$ and $f_{\bm{X}_n|\bm{X}_{n-1}}\big(\bm{x}_{n}|\bm{x}_{n-1}; \paramVec\big)$ rather than needing to evaluate these densities.
% Plug-and-play algorithms include Bayesian approaches like ABC and PMCMC \cite{toni09,andrieu10}, but here we use algorithms that enable maximum likelihood estimation.
% To our knowledge, the only plug-and-play methods that can be used to maximize the likelihood for POMP models of this complexity are iterated filtering algorithms \cite{ionides15}, which modify the well-known \emph{particle filter} \cite{arulampalam02}.
% The particle filter is often used to obtain values of latent states, but can also be used to obtain likelihood estimates of a POMP model.
% Iterated filtering algorithms such as IF2 \cite{ionides15} extend the particle filter by performing a random walk for each parameter and particle;
% these perturbations are carried out iteratively over multiple filtering operations, using the collection of parameters from the previous filtering pass as the parameter initialization for the next iteration, and decreasing the random walk variance at each step.

@@

If a mechanistic model including a feature (such as a representation of a mechanism, or the inclusion of a covariate) fits better than mechanistic models without that feature, and also has competitive fit compared to associative benchmarks, this may be taken as evidence supporting the scientific relevance of the feature.
As for any analysis of observational data, we must be alert to the possibility of confounding.
For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled or unmeasured covariate.
For a mechanism, the model feature could in principle explain the data by helping to account for some different unmodeled phenomenon.
The statistical evidence of a trend in transmission rate in this model could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from collinear covariates.
Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate, resulting in confounded mechanisms.
The robust statistical conclusion is that a model which allows for change fits better than one which does not---we argue that a decreasing transmission rate is a plausible way to explain this, but the incidence data themselves do not provide enough information to pin down the mechanism.

@@

Another approach is to compare a quantitative measure of the model fit (such as MSE, predictive accuracy, or model likelihood) among all proposed models.
These comparisons, which provide insight into how each model performs relative to the others, are quite common \cite{rinaldo12,sallah17}.
To calibrate relative measures of fit, it is useful to compare against a model that has well-understood statistical ability to fit data, and we call this model a {\it benchmark}.
Standard statistical models, interpreted as associative models without requiring any mechanistic interpretation of their parameters, provide suitable benchmarks.
Examples include linear regression, auto regressive moving average time series models, or even independent and identically distributed measurements.
Benchmarks enable us to evaluate the goodness of fit that can be expected of a suitable mechanistic model.

@@

In some cases, a possible benchmark model could be a generally accepted mechanistic model, but often no such model is available.
Because of this, we use a simple negative binomial model with an auto regressive mean as our associative benchmark; this model is described in \myeqref{eq:AR_NB}.
\begin{eqnarray}
\label{eq:AR_NB}
Y_n|Y_{n - 1} \sim \text{NB}\big(\NBintercept + \NBar Y_{n-1}, \NBsize \big),
\end{eqnarray}
where $\mathrm{E}\big(Y_n|Y_{n - 1}\big) = \NBintercept + \NBar Y_{n-1}$, and $\Var\big(Y_n|Y_{n - 1}\big) = \mathrm{E}\big(Y_n|Y_{n - 1}\big) + \mathrm{E}\big(Y_n|Y_{n - 1}\big)^2 / \NBsize$.
In the case where the case counts are large, an alternative benchmark recommended by He et al.~\cite{he10} is a log-linear Gaussian ARMA model;
the theory and practice of ARMA models is well developed, and these linear models are appropriate on a log scale due to the exponential growth and decay characteristic of biological dynamics.
We chose to use the auto regressive negative binomial model, however, because the large number of weeks with zero recorded cholera cases in department level data makes a benchmark based on a continuous distribution problematic.
Likelihoods of Models~1--3 and their respective ARMA benchmark models are provided in Table~\ref{tab:likes}.

@@

Our case study provides an example of this in the parameter estimate for the duration of natural immunity due to cholera infection, $\muRS^{-1}$.
Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],3)} \, \mathrm{yr}$, suggesting that individuals have effectively permanent immunity to cholera once infected.
To interpret this result, we bear in mind that the data ranged from 2010-2019, and therefore estimates of immunity longer than $10\, \mathrm{yr}$---the upper end of previous estimates of natural immunity \cite{king08}---effectively result in the same model dynamics.
The depletion of susceptible individuals may also be attributed to confounding mechanisms---such as localized vaccination programs and non-pharmaceutical interventions that reduce cholera transmission \cite{trevisin22, rebaudet21}---that were not accounted for in the model.
Perhaps the best interpretation of the estimated parameter, then, is that under the modeling framework that was used, the model most adequately describes the observed data by having a steady decrease in the number of susceptible individuals.
The weak statistical fit of Model~2 compared to a log-linear benchmark (see Table~\ref{tab:likes}) cautions us against drawing quantitative conclusions from this model.
A model that has a poor statistical fit may nevertheless provide a useful conceptual framework for thinking about the system under investigation.
However, a claim that the model has been validated against data should be reserved for situations where the model provides a statistical fit that is competitive against alternative explanations.

@@

@@

Recent information about a dynamic system should be more relevant for a forecast than older information.
This assertion may seem self-evident, but it is not the case for deterministic models, for which the initial conditions together with the parameters are sufficient for forecasting, and so recent data do not have special importance.
Epidemiological forecasts based on deterministic models are not uncommon in practice, despite their limitations \cite{king15}.
That may explain why Lee et al.~\cite{lee20} chose to obtain forecasts by simulating the calibrated models forward from initial conditions.
We obtain forecasts by simulating future values using latent states that are consistent with the most recent observations; this is done by drawing latent states at the last observation time ($t_N$) from the filtering distribution $f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$.
The decision not to do this partially explains the unsuccessful forecasts of Lee et al.~\cite{lee20}: their Table~S7 shows that the subset of their simulations which were consistent with observing zero cases in 2019 also accurately predicted the prolonged absence of detected cholera.
